---
title: Linear reg, and XGBoost
subtitle: Week 4 (con'd) - 02
date: 2025-06-09
format: 
    html:
        toc: true
        embed-resources: true
        warning: false
        tbl-cap-location: bottom
---

# Linear model – Regression & Non-linear model – XGBoost

```{r}
#| output: false
#| warning: false
library(reticulate)
library(readr)
library(tidyr)
library(knitr)
library(arrow)
library(tidymodels)
library(regclass)
library(xgboost)
set.seed(512) # Set seed to ensure code result reproducibility for randomization
```

## Data Import

```{r}
ohe_unimputed_train <- read_feather('../../dssg-2025-mentor-canada/Data/ohe_unimputed_train.feather')
ohe_unimputed_train <- ohe_unimputed_train %>% mutate_all(~as.numeric(as.character(.))) |>
                             mutate(any_mentor = ifelse(QS2_3_PRESENCEOFM_Yes == 1 |QS2_9_PRESENCEOFA_Yes == 1, 1, 0), 
                                    not_low_ses = ifelse(QS3_4_LIFEEVENTS1_11_11_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_16_16_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_18_18_X1 == 0, 1, 0)) 
data_unnorm <- # read_feather("../../dssg-2025-mentor-canada/Data/ohe_umimputed_all_data.feather") |>
read_csv("../../dssg-2025-mentor-canada/Data/faiss_knn_imputed_all_data.csv") |>
     mutate_all(~as.numeric(as.character(.))) |>
     mutate(any_mentor = ifelse(QS2_3_PRESENCEOFM_Yes == 1 |QS2_9_PRESENCEOFA_Yes == 1, 1, 0), 
                                    not_low_ses = ifelse(QS3_4_LIFEEVENTS1_11_11_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_16_16_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_18_18_X1 == 0, 1, 0),
                                     low_ses = ifelse(QS3_4_LIFEEVENTS1_11_11_X1 == 1 |
                                                                    QS3_4_LIFEEVENTS1_16_16_X1 == 1 |
                                                                    QS3_4_LIFEEVENTS1_18_18_X1 == 1, 1, 0))|>
                            filter(QS1_28_EMPLOYMENT_calculated < 542000)

mean_mode_imputed_train <- read_feather("../../dssg-2025-mentor-canada/Data/mean_mode_imputed_train.feather")

# knn_imputed_train <- read_csv('../../dssg-2025-mentor-canada/Data/faiss_knn_imputed_dataset.csv')
knn_imputed_train <- read_csv('../../dssg-2025-mentor-canada/Data/faiss_tuned_knn_imputed_train.csv')|>
                             mutate(any_mentor = ifelse(QS2_3_PRESENCEOFM_Yes == 1 |QS2_9_PRESENCEOFA_Yes == 1, 1, 0), 
                                    not_low_ses = ifelse(QS3_4_LIFEEVENTS1_11_11_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_16_16_X1 == 0 |
                                            QS3_4_LIFEEVENTS1_18_18_X1 == 0, 1, 0))

```

## Model fitting

### Linear model - Linear Regression (Ordinary Least Square Reg)

```{r}
#| echo: false
#| output: false
#| warning: false
mentor_ses_no_na_unimputed <- ohe_unimputed_train |>
                            select(QS1_28_EMPLOYMENT_calculated, QS3_4_LIFEEVENTS1_11_11_X1,QS3_4_LIFEEVENTS1_16_16_X1,
                                   QS3_4_LIFEEVENTS1_18_18_X1,QS2_3_PRESENCEOFM_Yes,QS2_9_PRESENCEOFA_Yes,
                                   QS2_6_MENTOREXPER, QS1_1_AGE) |>
                            na.omit()
subset <- ohe_unimputed_train |>
select(QS1_28_EMPLOYMENT_calculated, QS3_4_LIFEEVENTS1_11_11_X1,QS3_4_LIFEEVENTS1_16_16_X1,
                                   QS3_4_LIFEEVENTS1_18_18_X1,QS2_3_PRESENCEOFM_Yes,QS2_9_PRESENCEOFA_Yes,
                                   QS2_6_MENTOREXPER, QS1_1_AGE)
colSums(is.na(subset))

```

```{r}
#---unimputed
# Use this:
model <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            QS1_1_AGE , 
            data = ohe_unimputed_train)
summary(model)

# ver 2

model <- lm(QS1_28_EMPLOYMENT_calculated ~  not_low_ses * any_mentor +
                                           # QS2_6_MENTOREXPER + 
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model) # those who had a mentor faired worse if they were in low ses--if one is not in low ses and had a mentor, income seem to be positively associated
# having a mentor helps those who were in higher SES *****


model <- lm(QS1_28_EMPLOYMENT_calculated ~  low_ses * any_mentor +
                                           # QS2_6_MENTOREXPER + 
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model)

# ver 3
model <- lm(QS1_28_EMPLOYMENT_calculated ~ QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS1_1_AGE + 
                                            QS2_3_PRESENCEOFM_Yes*QS2_9_PRESENCEOFA_Yes, 
            data = data_unnorm)
summary(model)

ggplot(data_unnorm, aes(x = QS1_28_EMPLOYMENT_calculated, y = QS1_28_EMPLOYMENT_calculated)) + geom_point()

#----knn imputed
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_imputed)

#----logistic unimputed
lg_model_unimputed <- glm(QS2_9_PRESENCEOFA_Yes ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                         QS3_4_LIFEEVENTS1_16_16_X1 + 
                                         QS3_4_LIFEEVENTS1_18_18_X1 + 
                                         QS2_3_PRESENCEOFM_Yes + 
                                         QS1_1_AGE, 
            data = mean_mode_imputed_train,
            family = binomial)
summary(lg_model_unimputed)

#----logistic knn imputed
lg_model <- glm(QS2_9_PRESENCEOFA_Yes ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                         QS3_4_LIFEEVENTS1_16_16_X1 + 
                                         QS3_4_LIFEEVENTS1_18_18_X1 + 
                                         QS2_3_PRESENCEOFM_Yes + 
                                         QS1_1_AGE, 
            data = knn_imputed_train,
            family = binomial)
summary(lg_model)
```

> -   Low R-squared suggests that the predictors, as specified, have limited practical/explanatory power for income.

### Running linear regression KNN imputed data:

```{r}
# Use this:
# ver 1:
model_knn_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes + 
                                            QS2_9_PRESENCEOFA_Yes + 
                                           # QS2_6_MENTOREXPER + 
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_knn_imputed)

# ----
# ver 2
model_knn_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  not_low_ses + 
                                            any_mentor +
                                           # QS2_6_MENTOREXPER + 
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_knn_imputed)

lg_model_knn_imputed <- glm(QS2_9_PRESENCEOFA_Yes ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                         QS3_4_LIFEEVENTS1_16_16_X1 + 
                                         QS3_4_LIFEEVENTS1_18_18_X1 + 
                                         QS2_3_PRESENCEOFM_Yes + 
                                         QS1_1_AGE, 
            data = knn_imputed_train,
            family = binomial)
summary(lg_model_knn_imputed)
```

### Year since completing higher education (instead of age)

```{r}
model_knn_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes + 
                                            QS2_9_PRESENCEOFA_Yes + 
                                            QS1_1_AGE +
                        QS2_3_PRESENCEOFM_Yes*QS3_4_LIFEEVENTS1_11_11_X1 +
                        QS2_3_PRESENCEOFM_Yes*QS3_4_LIFEEVENTS1_16_16_X1 +
                        QS2_3_PRESENCEOFM_Yes*QS3_4_LIFEEVENTS1_18_18_X1 +
                        QS2_9_PRESENCEOFA_Yes*QS3_4_LIFEEVENTS1_11_11_X1 +
                        QS2_9_PRESENCEOFA_Yes*QS3_4_LIFEEVENTS1_16_16_X1 +
                        QS2_3_PRESENCEOFM_Yes*QS3_4_LIFEEVENTS1_18_18_X1 
                       , 
            data = knn_imputed_train)
summary(model_knn_imputed)

mentor_alone <- lm(QS1_28_EMPLOYMENT_calculated ~ 
                                            QS2_3_PRESENCEOFM_Yes + 
                                            QS2_9_PRESENCEOFA_Yes + 
                                            QS1_1_AGE, 
            data = knn_imputed_train)
summary(mentor_alone)

```

```{r}
#----knn imputed income

knn_imputed_train <- knn_imputed_train |>
                      mutate(mentor_6to18 = ifelse(QS2_3_PRESENCEOFM_Yes == 1 | QS2_9_PRESENCEOFA_Yes == 1, 1, 0))
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            QS2_37_HELPFULNESS+
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_imputed)

# Use this:
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            # QS2_3_PRESENCEOFM_Yes +
                                            # QS2_9_PRESENCEOFA_Yes + 
                                            QS2_37_HELPFULNESS+
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_imputed)

# ver 2:
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            # QS2_3_PRESENCEOFM_Yes +
                                            # QS2_9_PRESENCEOFA_Yes + 
                                            QS2_37_HELPFULNESS+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)

# ver 2:
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~ low_ses + 
                                            # QS2_3_PRESENCEOFM_Yes +
                                            # QS2_9_PRESENCEOFA_Yes + 
                                            
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)

model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)                                                 # *****

model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            # QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = knn_imputed_train)
summary(model_imputed)

# interaction 1
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            # QS2_6_MENTOREXPER+
                                            QS1_1_AGE +
                                            QS3_4_LIFEEVENTS1_11_11_X1*QS2_3_PRESENCEOFM_Yes + 
                                            QS3_4_LIFEEVENTS1_16_16_X1*QS2_3_PRESENCEOFM_Yes + 
                                            QS3_4_LIFEEVENTS1_18_18_X1*QS2_3_PRESENCEOFM_Yes, 
            data = knn_imputed_train)
summary(model_imputed)

# interaction 2
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            # QS2_3_PRESENCEOFM_Yes +
                                            # QS2_9_PRESENCEOFA_Yes + 
                                            QS2_37_HELPFULNESS+
                                            QS1_1_AGE +
                                            QS3_4_LIFEEVENTS1_11_11_X1*QS2_37_HELPFULNESS + 
                                            QS3_4_LIFEEVENTS1_16_16_X1*QS2_37_HELPFULNESS + 
                                            QS3_4_LIFEEVENTS1_18_18_X1*QS2_37_HELPFULNESS, 
            data = knn_imputed_train)
summary(model_imputed)
```

```{r}
# unimputed
model_imputed_unimputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            # QS2_3_PRESENCEOFM_Yes +
                                            # QS2_9_PRESENCEOFA_Yes + 
                                            QS2_37_HELPFULNESS+
                                            QS1_1_AGE +
                                            QS3_4_LIFEEVENTS1_11_11_X1*QS2_37_HELPFULNESS + 
                                            QS3_4_LIFEEVENTS1_16_16_X1*QS2_37_HELPFULNESS + 
                                            QS3_4_LIFEEVENTS1_18_18_X1*QS2_37_HELPFULNESS, 
            data = data_unnorm)
summary(model_imputed_unimputed)
```

#### Multicollinearity:

```{r}
VIF(model)
```

## Mentor Influence on Annual Income

```{r}
model_mentors_influence <- lm(QS1_28_EMPLOYMENT_calculated ~  QS2_36_INFLUENCE1_1_1 + 
                                                              QS2_36_INFLUENCE1_2_2 + 
                                                              QS2_36_INFLUENCE1_3_3 + 
                                                              QS2_36_INFLUENCE1_4_4 + 
                                                              QS2_36_INFLUENCE1_5_5 + 
                                                              QS2_36_INFLUENCE1_6_6 + 
                                                              QS2_36_INFLUENCE1_7_7 + 
                                                              QS2_36_INFLUENCE1_8_8 + 
                                                              QS2_36_INFLUENCE1_9_9 + QS1_1_AGE, 

                                data = mean_mode_imputed_train)
summary(model_mentors_influence)
```

```{r}
model_mentors_influence_knn_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS2_36_INFLUENCE1_1_1 + 
                                                              QS2_36_INFLUENCE1_2_2 + 
                                                              QS2_36_INFLUENCE1_3_3 + 
                                                              QS2_36_INFLUENCE1_4_4 + 
                                                              QS2_36_INFLUENCE1_5_5 + 
                                                              QS2_36_INFLUENCE1_6_6 + 
                                                              QS2_36_INFLUENCE1_7_7 + 
                                                              QS2_36_INFLUENCE1_8_8 + 
                                                              QS2_36_INFLUENCE1_9_9 + QS1_1_AGE, 

                                data = knn_imputed_train)
summary(model_mentors_influence_knn_imputed)
```

## Model specification (Lasso regression with L1 regularization with `mixture = 1`)

(Workflow with tidymodels–for now has echo: false)

```{r}
#| echo: false
# ridge_spec <- linear_reg(penalty = tune(), mixture = 0) |> 
# set_mode("regression") |> 
# set_engine("glmnet")
```

```{r}
#| echo: false

# fit <- linear_reg(penalty = 1) |>
#         set_engine("glmnet") |>
#         fit(QS1_28_EMPLOYMENT_calculated ~ ., data = mean_mode_imputed_train)

# lm_wkflw <- workflow() |>
#         add_recipe(recipe_to_fit) |>
#         add_model(lm_spec)
#         fit(lm_wkflw, data = mean_mode_imputed_train)
```

#### Define cross-validation (CV) folds for tuning & grid:

```{r}
#| echo: false

# cv_folds <- vfold_cv(imputed_train_fit, v = 10)
# grid <- grid_regular(penalty(range = c(-3, 3)), levels = 20)
```

### `Tidymodels` Workflow

```{r}
#| echo: false

#ridge_wkflw <- workflow() |>       
#add_recipe(recipe_to_fit) |>     
#add_model(ridge_spec)
```

### Tuning Regularization Penalty

```{r}
#| echo: false

# ridge_tune <- tune_grid(ridge_wkflw, resamples = cv_folds, grid = grid, metrics = metric_set(rmse))
```

```{r}
#| echo: false

# Fit Ridge model (alpha = 0)
# View model summary

```

```{r}
#| echo: false

  # Perform cross-validation to select optimal lambda
# x_train <- select(mean_mode_imputed_train, -QS1_28_EMPLOYMENT_calculated)
# y_train <- select(mean_mode_imputed_train, QS1_28_EMPLOYMENT_calculated)

# # Optimal lambda
# best_lambda <- cv_ridge$lambda.min

# # Plot cross-validation results
# plot(cv_ridge)
```

```{r}
#| echo: false

#   ses_youth <- select(youth, QS3_4_LIFEEVENTS1_11_11, QS3_4_LIFEEVENTS1_16_16, QS3_4_LIFEEVENTS1_18_18, QS1_28_EMPLOYMENT_calculated)
# rec <- recipe(QS1_28_EMPLOYMENT_calculated~QS3_4_LIFEEVENTS1_11_11 + QS3_4_LIFEEVENTS1_16_16 + QS3_4_LIFEEVENTS1_18_18, data = ses_youth)
# dummies_recipe <- rec |> step_dummy(all_predictors(), one_hot = TRUE)
# dummies_data <- dummies_recipe |> prep() |> bake(new_data = NULL)
# dummies_data
```

## XGB

### XGB model 1:

#### All 3 SES indicators, mentor experiences, age, and year of higher education completion.

**(Mean mode imputed)**:

```{r}
#| output: false
x <- model.matrix(~ QS3_4_LIFEEVENTS1_16_16_X1 + QS2_3_PRESENCEOFM_Yes + 
                  QS2_9_PRESENCEOFA_Yes + QS2_6_MENTOREXPER + 
                  QS1_1_AGE + QS1_23_YEARCOMPLE, 
                  data = mean_mode_imputed_train)[, -1]
y <- log(mean_mode_imputed_train$QS1_28_EMPLOYMENT_calculated + 1)
dmatrix <- xgb.DMatrix(x, label = y)
xgb_model <- xgboost(data = dmatrix, nrounds = 100, objective = "reg:squarederror", 
                     params = list(max_depth = 4, eta = 0.1))
xgb.importance(model = xgb_model)

xgb.plot.importance(xgb.importance(model = xgb_model))

```

```{r}
#| output: false
xgb_importance_df <- xgb.importance(model = xgb_model)

xgb_cv <- xgb.cv(params = list(max_depth = 4, eta = 0.1), data = dmatrix, nrounds = 200, nfold = 5)
```

```{r}
xgb.plot.importance(xgb.importance(model = xgb_model))
kable(xgb_importance_df)
```

**(KNN imputed)**:

```{r}
#| output: false
x_knn_imputed <- model.matrix(~ QS3_4_LIFEEVENTS1_16_16_X1 + QS2_3_PRESENCEOFM_Yes + 
                  QS2_9_PRESENCEOFA_Yes + QS2_6_MENTOREXPER + 
                  QS1_1_AGE + QS1_23_YEARCOMPLE, 
                  data = knn_imputed_train)[, -1]
y_knn_imputed <- log(knn_imputed_train$QS1_28_EMPLOYMENT_calculated + 1)
dmatrix_knn_imputed <- xgb.DMatrix(x_knn_imputed, label = y_knn_imputed)
xgb_model_knn_imputed <- xgboost(data = dmatrix_knn_imputed, nrounds = 100, objective = "reg:squarederror", 
                     params = list(max_depth = 4, eta = 0.1))

xgb.importance(model = xgb_model_knn_imputed)

xgb.plot.importance(xgb.importance(model = xgb_model_knn_imputed))

```

```{r}
#| output: false
xgb_importance_knn_imputed_df <- xgb.importance(model = xgb_model_knn_imputed)

xgb_cv_knn_imputed <- xgb.cv(params = list(max_depth = 4, eta = 0.1), data = dmatrix, nrounds = 200, nfold = 5)
```

```{r}
xgb.plot.importance(xgb.importance(model = xgb_model_knn_imputed))
kable(xgb_importance_knn_imputed_df)
```

### XGB model 2:

#### All 3 SES indicators, mentor experiences, and age.

```{r}
#| output: false
x <- model.matrix(~ QS3_4_LIFEEVENTS1_11_11_X1 + QS3_4_LIFEEVENTS1_16_16_X1 +  QS3_4_LIFEEVENTS1_18_18_X1 + QS2_3_PRESENCEOFM_Yes + QS2_9_PRESENCEOFA_Yes + QS2_6_MENTOREXPER + QS1_1_AGE, data = mean_mode_imputed_train)[, -1]

y <- log(mean_mode_imputed_train$QS1_28_EMPLOYMENT_calculated + 1)
dmatrix <- xgb.DMatrix(x, label = y)
xgb_model <- xgboost(data = dmatrix, nrounds = 100, objective = "reg:squarederror", 
params = list(max_depth = 4, eta = 0.1))

xgb_importance_df <- xgb.importance(model = xgb_model)

xgb_cv <- xgb.cv(params = list(max_depth = 4, eta = 0.1), data = dmatrix, nrounds = 200, nfold = 5)
```

```{r}
xgb.plot.importance(xgb.importance(model = xgb_model))
write_csv(xgb_importance_df, "outputs/tables/week-04/02-xgb-importance.csv")
kable(xgb_importance_df)
```

**(KNN imputed)**:

```{r}
#| output: false
x_knn_imputed <- model.matrix(~ QS3_4_LIFEEVENTS1_11_11_X1 + QS3_4_LIFEEVENTS1_16_16_X1 +  QS3_4_LIFEEVENTS1_18_18_X1 + QS2_3_PRESENCEOFM_Yes + QS2_9_PRESENCEOFA_Yes + QS2_6_MENTOREXPER + QS1_1_AGE, 
                  data = knn_imputed_train)[, -1]
y_knn_imputed <- log(knn_imputed_train$QS1_28_EMPLOYMENT_calculated + 1)
dmatrix_knn_imputed <- xgb.DMatrix(x_knn_imputed, label = y_knn_imputed)
xgb_model_knn_imputed <- xgboost(data = dmatrix_knn_imputed, nrounds = 100, objective = "reg:squarederror", 
                     params = list(max_depth = 4, eta = 0.1))
xgb.importance(model = xgb_model_knn_imputed)

xgb.plot.importance(xgb.importance(model = xgb_model_knn_imputed))

```

```{r}
#| output: false
xgb_importance_knn_imputed_df <- xgb.importance(model = xgb_model_knn_imputed)

xgb_cv_knn_imputed <- xgb.cv(params = list(max_depth = 4, eta = 0.1), data = dmatrix, nrounds = 200, nfold = 5)
```

```{r}
xgb.ggplot.importance(xgb.importance(model = xgb_model_knn_imputed), rel_to_first = TRUE) +
         scale_fill_brewer(palette = 'Accent') +
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14),
        legend.text = element_text(size = 11)) 

kable(xgb_importance_knn_imputed_df)
```

### Mentor helpfulness instead of Mentor experience:

```{r}
#| output: false
x_knn_imputed <- model.matrix(~ QS3_4_LIFEEVENTS1_11_11_X1 + QS3_4_LIFEEVENTS1_16_16_X1 +  QS3_4_LIFEEVENTS1_18_18_X1 + QS2_37_HELPFULNESS + QS1_1_AGE, 
                  data = knn_imputed_train)[, -1]
y_knn_imputed <- log(knn_imputed_train$QS1_28_EMPLOYMENT_calculated + 1)
dmatrix_knn_imputed <- xgb.DMatrix(x_knn_imputed, label = y_knn_imputed)
xgb_model_knn_imputed <- xgboost(data = dmatrix_knn_imputed, nrounds = 100, objective = "reg:squarederror", 
                     params = list(max_depth = 4, eta = 0.1))
xgb.importance(model = xgb_model_knn_imputed)

xgb.plot.importance(xgb.importance(model = xgb_model_knn_imputed))
```

```{r}
#| output: false
xgb_importance_knn_imputed_df <- xgb.importance(model = xgb_model_knn_imputed)

xgb_cv_knn_imputed <- xgb.cv(params = list(max_depth = 4, eta = 0.1), data = dmatrix, nrounds = 200, nfold = 5)
```

```{r}
xgb.ggplot.importance(xgb.importance(model = xgb_model_knn_imputed), rel_to_first = TRUE) +
         scale_fill_brewer(palette = 'Accent') +
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14),
        legend.text = element_text(size = 11)) 

kable(xgb_importance_knn_imputed_df)
```

------------------------------------------------------------------------

## Predicting mental health:

> 

```{r}
model_imputed <- lm(QS4_9_MENTALHEALTH ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_3_PRESENCEOFM_Yes +
                                            QS2_9_PRESENCEOFA_Yes + 
                                            # QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed) # *****

model_imputed <- lm(QS4_9_MENTALHEALTH ~  not_low_ses *  QS2_6_MENTOREXPER +
                                          
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed) # ***** # model 2

model_imputed <- lm(QS4_9_MENTALHEALTH ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed) # *****

```

> Full low SES indicators interpretation:
>
> The linear regression model evaluates the impact of specific low socioeconomic status (SES) indicators in youth—family financial assistance (QS3_4_LIFEEVENTS1_11_11_X1), working to support family (QS3_4_LIFEEVENTS1_16_16_X1), and food bank use (QS3_4_LIFEEVENTS1_18_18_X1)—along with mentorship presence in early (ages 6–11, **QS2_3_PRESENCEOFM_Yes**) and later youth (ages 12–18, QS2_9_PRESENCEOFA_Yes), and current age on adult mental health (QS4_9_MENTALHEALTH).
>
> Having to work to support the family in youth is associated with better mental health (β = 0.1786, p = 0.015), while food bank use is linked to worse mental health (β = -0.1895, p = 0.023). A mentor’s presence from ages 6–11 significantly improves mental health (β = 0.2673, p \< 0.001). Financial assistance, later mentorship (ages 12–18), and age are not significant predictors.

> If using `not_low_ses`:
>
> The linear regression model assesses how adult mental health (QS4_9_MENTALHEALTH) is influenced by youth socioeconomic status (SES), the presence of a mentor during ages 6–11 and 12–18, and current age.
>
> > Having a mentor from ages 6–11 (QS2_3_PRESENCEOFM_Yes) is associated with better mental health (β = 0.2572, p \< 0.001), increasing the mental health rating by about 0.26 points. Not having low SES in youth (not_low_ses) is linked to **slightly worse** mental health (β = -0.2675, p = 0.034).
> >
> > The presence of a mentor from ages 12–18 and age do **not** significantly predict mental health.

```{r}

model_imputed <- lm(QS4_9_MENTALHEALTH ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_37_HELPFULNESS +
                                            # QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed) # *****


```

```{r}
# Given that the participant has mentor:
model_imputed <- lm(QS4_9_MENTALHEALTH ~  not_low_ses * QS2_6_MENTOREXPER +
                                            QS2_6_MENTOREXPER * QS2_37_HELPFULNESS+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed) # *****
```

> The linear regression model examines the relationship between adult mental health (**QS4_9_MENTALHEALTH**) and youth socioeconomic status (SES), mentor experience quality, mentor helpfulness, and age. Having more positive mentor experiences from ages 6–18 (**QS2_6_MENTOREXPER**) is associated with better mental health (β = 0.4896, p = 0.016), with each unit increase in positive mentor experience raising mental health ratings by about 0.49 points.
>
> Older age is linked to slightly worse mental health (β = -0.0152, p = 0.027), with each year reducing the rating by 0.015 points.
>
> Neither SES, mentor helpfulness, nor their interactions significantly predict mental health. The model explains only 2.9% of the variance in mental health (R² = 0.029)

## Previous models predicting income:

```{r}
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)   

model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  not_low_ses*  QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)  

model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  not_low_ses*  QS2_37_HELPFULNESS+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)  
# *****
```

```{r}
# ver 2

model <- lm(QS1_28_EMPLOYMENT_calculated ~  not_low_ses * any_mentor +
                                           # QS2_6_MENTOREXPER + 
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model) # those who had a mentor faired worse if they were in low ses--if one is not in low ses and had a mentor, income seem to be positively associated
# having a mentor helps those who were in higher SES *****


```

> > for those from low SES backgrounds, having a mentor is associated with a \$19,218 decrease in annual income, which is statistically significant but **counterintuitive** and may warrant further investigation (e.g., type of mentorship or other confounding factors).
> >
> > The interaction term (\$not_low_ses:any_mentor\$, β = 20751.1, p = 0.035) indicates that for individuals from non-low SES backgrounds (not_low_ses = 1), the effect of having a mentor is positive, increasing income by an additional \$20,751 compared to those from low SES with a mentor.

```{r}
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_6_MENTOREXPER+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)                                                 # *****
```

> 

```{r}
model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  QS3_4_LIFEEVENTS1_11_11_X1 + 
                                            QS3_4_LIFEEVENTS1_16_16_X1 + 
                                            QS3_4_LIFEEVENTS1_18_18_X1 + 
                                            QS2_34_SUPPORTS1_9_9+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)   

model_imputed <- lm(QS1_28_EMPLOYMENT_calculated ~  low_ses * QS4_3_CAREERPLANNIN1_3_3+
                                            QS1_1_AGE , 
            data = data_unnorm)
summary(model_imputed)   
```

## Mediation Analysis

```{r}
library(mediation)
data<-data_unnorm
 # Prepare data
 data$low_ses <- as.integer(data$QS3_4_LIFEEVENTS1_11_11_X1 | data$QS3_4_LIFEEVENTS1_16_16_X1 | data$QS3_4_LIFEEVENTS1_18_18_X1)
 data$mentorship <- data$QS2_3_PRESENCEOFM_Yes  # Binary mentorship variable
 data$adult_ses <- log(data$QS1_28_EMPLOYMENT_calculated + 1)  # Log-transform income, assuming continuous
 
 # Step 1: Path a (SES -> Mentorship)
 model_m <- glm(mentorship ~ low_ses + QS1_1_AGE + QS2_9_PRESENCEOFA_Yes, 
                family = binomial, data = data)
 
 # Step 2: Path b and c' (Mentorship and SES -> Adult SES)
 model_y <- lm(adult_ses ~ low_ses + mentorship + QS1_1_AGE + QS2_9_PRESENCEOFA_Yes, 
               data = data)
 
 # Step 3: Mediation analysis
 med_model <- mediate(model_m, model_y, treat = "low_ses", mediator = "mentorship", 
                     boot = TRUE, sims = 1000)
 
 # Summary of results
 summary(med_model)

```

```{r}
 # Prepare data
 data$low_ses <- as.integer(data$QS3_4_LIFEEVENTS1_11_11_X1 | data$QS3_4_LIFEEVENTS1_16_16_X1 | data$QS3_4_LIFEEVENTS1_18_18_X1)
 data$mentorship <- data$QS2_3_PRESENCEOFM_Yes  # Binary mentorship variable
 data$adult_mental_health <- data$QS4_9_MENTALHEALTH  # Log-transform income, assuming continuous
 
 # Step 1: Path a (SES -> Mentorship)
 model_m <- glm(mentorship ~ low_ses + QS1_1_AGE + QS2_9_PRESENCEOFA_Yes, 
                family = binomial, data = data)
 
 # Step 2: Path b and c' (Mentorship and SES -> Adult SES)
 model_y <- lm(adult_mental_health ~ low_ses + mentorship + QS1_1_AGE + QS2_9_PRESENCEOFA_Yes, 
               data = data)
 
 # Step 3: Mediation analysis
 med_model <- mediate(model_m, model_y, treat = "low_ses", mediator = "mentorship", 
                     boot = TRUE, sims = 1000)
 
 # Summary of results
 summary(med_model)

```