---
title: Descriptive Data Statistics
format: 
    html:
        toc: true
        embed-resources: true
        warning: false
---

## Python: Data Cleaning & Feature Engineering

### Library imports

```{r}
library(reticulate)
library(here)
py_config()
```

```{python}
import pandas as pd
import os
from siuba import _, group_by, summarize, filter, select, mutate, arrange, count
# report_directory = "quarto/" # setting working directory into the quarto folder to ensure quarto/code execution consistency
# os.chdir(report_directory)
```

### Data Import

```{python}
file_path = os.path.join("..", "..", "dssg-2025-mentor-canada", "Data", "Data_2020-Youth-Survey.csv")

youth = pd.read_csv(file_path)
```

### Preview the first 6 rows and 7 columns.

```{python}
youth.iloc[:6,:8]
```

```{python}
#| warning: false
#| output: asis
youth.info()
```

```{python}
print(youth.shape)

null_counts = youth.isnull().sum()
null_counts = null_counts[null_counts > 0]
```

```{python}
null_counts = null_counts.reset_index()
null_counts
```

### Reversing one-hot encoding for the gender identity column:

```{python}
gender_cols = youth.loc[:,'QS1_9_GENDER1_1_1':'QS1_9_GENDER1_6_6']
youth['QS1_9_gender'] = gender_cols.idxmax(axis = 1)
youth['QS1_9_gender'].head().reset_index()
```

> QS1_28_EMPLOYMENT_calculated, gender_cols

### Create estimated total year income column:

> (Feature engineering)

```{python}
youth['yearly_from_month'] = youth['Month_income'].fillna(0) * 12
youth['yearly_from_seimimonth'] = youth['Semimonth_income'].fillna(0) * 24
youth['yearly_from_biweek'] = youth['Biweek_income'].fillna(0)*26
youth['yearly_from_week'] = youth['Week_income'].fillna(0) * 52

youth['total_yearly_income'] = (youth['yearly_from_month'] +
                                youth['yearly_from_seimimonth'] + 
                                youth['yearly_from_biweek'] + 
                                youth['yearly_from_week'])

```

#### Preview new `total_yearly_income` column:

```{python}
(youth['total_yearly_income']).head()
```

#### Save manipulated dataset to local as an intermediate processed dataset:

```{python}
youth.to_csv('../../../dssg-2025-mentor-canada/Data/intermediate.csv')
```

## R: Visualization

```{r}
#| warning: false
library(tidyverse)
library(knitr)
library(reticulate)
```

```{r}
#| warning: false
youth <- read_csv('../../../dssg-2025-mentor-canada/Data/intermediate.csv')
kable(head(youth))
```

### **Histograms to examine distributions**:

#### Total estimated yearly income:

```{r}
ggplot(youth, aes(x = total_yearly_income)) +
    geom_histogram() +
    scale_y_log10() +
    labs(title = "Distribution of total estimated yearly income",
    x = "Estimate Total Yearly Income ($)") +
    theme_minimal()
```

#### Frequency count of each gender identity (`QS1_9_gender`):

```{r}
ggplot(youth, aes(x = QS1_9_gender, fill = QS1_9_gender)) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Compare gender identity to estimated yearly total income (`total_yearly_income`):

```{r}
ggplot(youth, aes(x = QS1_9_gender, y = total_yearly_income, fill = QS1_9_gender)) +
    geom_bar(stat = "identity") +
    labs(x = "Gender Indentity", y = "Total estimated yearly income ($)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(youth, aes(x = QS1_9_gender, y = total_yearly_income, fill = QS1_9_gender)) +
    geom_boxplot() +
    labs(x = "Gender Indentity", y = "Total estimated yearly income ($)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
youth |>
filter(total_yearly_income < 29380) |>
ggplot( aes(x = QS1_9_gender, y = total_yearly_income, fill = QS1_9_gender)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(x = "Gender Indentity", y = "Total estimated yearly income ($)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Comparing further education (`QS1_21_FURTHEDUCA`) and current income:

```{r}
ggplot(youth, aes(x = QS1_21_FURTHEDUCA, y = total_yearly_income, fill = QS1_9_gender)) +
    geom_boxplot(outliers = FALSE) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> #### Remove outlier in `no` group by removing the observations with `total_yearly_income` \$29,380, the low-income for an individual person in 2024 in canada.

```{r}
youth |>
filter(total_yearly_income < 29380) |>
ggplot(aes(x = QS1_21_FURTHEDUCA, y = total_yearly_income, fill = QS1_9_gender)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Visualize Mental health outcome `QS4_9_MENTALHEALTH` (or `QS4_9_MENTALHEALTH_cat`) / `QS4_11_BELONGING` depending on early mentee experience (`QS2_3_PRESENCEOFM_cat` or `QS2_3_PRESENCEOFM`)

```{r}
youth |>
    filter(!is.na(QS4_9_MENTALHEALTH_cat)) |>
    drop_na(QS2_3_PRESENCEOFM_cat) |>
    ggplot(aes(x = QS2_3_PRESENCEOFM_cat, fill = QS4_9_MENTALHEALTH_cat)) +
    geom_bar(position = "dodge") +
    labs(x = "Early life mentor experience (age 6-11 years old)", fill = "Current Mental Health Rating")
```